No general trend for multiple dimensions of functional diversity in bird and mammal communities

### 2024-09-23 paper/GEB Submission/manuscript_GEB.Rmd
Minor spelling changes

We synthesize thousands of mammal and bird functional diversity time series to assess mean trends across communities.
We evaluate 1) whether or not there is a prevailing type of functional change across communities and if community characteristics explain variation in those trends, 2) hypotheses of change for individual trait axes, and 3) the implications of study-level trends for the ecology of those communities.
To address these goals, we present results for general trends that aggregate across all communities, category-specifictrends that aggregate across communities with similar characteristics, and study-level trends.

@@

We calculated yearly metrics of functional and species diversity for each time series.
Species-based metrics include species richness (*S*) and Jaccard similarity (*J*) as a measure of turnover.
Jaccard similarity was calculated relative to the first observed year for a time series.
A negative trend in *J* would therefore indicate decreasing similiarity.
We did not impose a correction for unobserved species as non-parametric estimators do not assign species identities to corrected richness values, and therefore could not be propagated to the functional diversity metrics.

@@

To test for category-specifictrends within and between different levels of taxa, biome, realm, and protection status we fit separate models with each of those factors added as a predictor interacting with time to the original model structure.
We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies. We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
All models in our analysis were fit using the *lme4* (v1.1-35.3) package in R (v4.3.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].

@@

category-specifictrends in turnover highlight the types of communities experiencing change in community composition. For example, we found significant turnover for most biomes, realms, and taxa with the exception of marine, tropical, and mammal studies. The lack of turnover in marine studies stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial [@blowes2019].
However, previous global estimates are dominated by fish communities which we exclude here and are likely driving the overall trend while disguising relative stasis in marine bird and mammal communities. We also found significant turnover for unprotected areas and no significant turnover in protected areas, indicating that in general unprotected communities are experiencing bigger shifts in community composition through time.

@@

The next largest group of studies (7 studies) were those with no species richness trend that showed a trend in at least one functional diversity metric. Based on a species approach alone, these communities apear to either be experiencing no change or some change in species identity, with unknown implications for conservation concern or ecological process. However, with the addition of a functional perspective we found multiple kinds of underlying functional change including increases and decreases in corrected functional richness and divergence and loss of corrected functional evenness. We therefore found evidence of both increases and decreases in the functional capacity of communities independent of species-based metrics, reinforcing the importance of moving beyond solely species-based approaches for biodiversity assessment [@hillebrand2017; @santini2017]. All but one of the studies in this group experienced significant turnover, which was also reflected in trends for multiple *CWM*'s, indicating that the overall functional space was shifting even when some characteristics of the functional space were maintained.

>>>>

We synthesize thousands of mammal and bird functional diversity time series to assess mean trends across communities.
We evaluate 1) whether or not there is a prevailing type of functional change across communities and if community characteristics explain variation in those trends, 2) hypotheses of change for individual trait axes, and 3) the implications of study-level trends for the ecology of those communities.
To address these goals, we present results for general trends that aggregate across all communities, category-specific trends that aggregate across communities with similar characteristics, and study-level trends.

@@

We calculated yearly metrics of functional and species diversity for each time series.
Species-based metrics include species richness (*S*) and Jaccard similarity (*J*) as a measure of turnover.
Jaccard similarity was calculated relative to the first observed year for a time series.
A negative trend in *J* would therefore indicate decreasing similarity.
We did not impose a correction for unobserved species as non-parametric estimators do not assign species identities to corrected richness values, and therefore could not be propagated to the functional diversity metrics.

@@

To test for category-specific trends within and between different levels of taxa, biome, realm, and protection status we fit separate models with each of those factors added as a predictor interacting with time to the original model structure.
We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies. We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
All models in our analysis were fit using the *lme4* (v1.1-35.3) package in R (v4.3.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].

@@

category-specific trends in turnover highlight the types of communities experiencing change in community composition. For example, we found significant turnover for most biomes, realms, and taxa with the exception of marine, tropical, and mammal studies. The lack of turnover in marine studies stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial [@blowes2019].
However, previous global estimates are dominated by fish communities which we exclude here and are likely driving the overall trend while disguising relative stasis in marine bird and mammal communities. We also found significant turnover for unprotected areas and no significant turnover in protected areas, indicating that in general unprotected communities are experiencing bigger shifts in community composition through time.

@@

The next largest group of studies (7 studies) were those with no species richness trend that showed a trend in at least one functional diversity metric. Based on a species approach alone, these communities appear to either be experiencing no change or some change in species identity, with unknown implications for conservation concern or ecological process. However, with the addition of a functional perspective we found multiple kinds of underlying functional change including increases and decreases in corrected functional richness and divergence and loss of corrected functional evenness. We therefore found evidence of both increases and decreases in the functional capacity of communities independent of species-based metrics, reinforcing the importance of moving beyond solely species-based approaches for biodiversity assessment [@hillebrand2017; @santini2017]. All but one of the studies in this group experienced significant turnover, which was also reflected in trends for multiple *CWM*'s, indicating that the overall functional space was shifting even when some characteristics of the functional space were maintained.
